---
title: "MAGMA_Celltyping: Getting started"
author: "<h4>Authors: <i>Brian M. Schilder, Alan Murphy, Julien Bryois & Nathan Skene</i></h4>"
date: "<h4>Updated: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i></h4>"
output:
  BiocStyle::html_document:
vignette: >
    %\VignetteIndexEntry{MAGMA_Celltyping} 
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=TRUE, message=FALSE}
library(MAGMA.Celltyping)
library(ggplot2)
library(dplyr)
```

# Intro 

`MAGMA.Celltyping` is a software package that facilitates conducting cell-type-specific enrichment tests on GWAS summary statistics. 

# Setup 

Specify where you want the large files to be downloaded to. 

*NOTE*: Make sure you change `storage_dir` to somewhere other than `tempdir()` 
if you want to make sure the results aren't deleted after this R session closes!

```{r, eval=FALSE}
storage_dir <- tempdir()
```


# Prepare data

## GWAS

- We need to have a summary statistics file to analyse as input.  
- As an example, you can download UK Biobank summary statistics 
for 'fluid_intelligence' using `get_example_gwas()`.

Here we provide a 
pre-munged version of the above file.

### Munging 

Our lab have created [`MungeSumstats`](https://github.com/neurogenomics/MungeSumstats), a robust Bioconductor package for formatting multiple types
of summary statistics files. We highly recommend processing your GWAS summary statistics with
`MungeSumstats` before continuing.
See the *full_workflow* vignette for more details.

The minimum info needed *after* munging is:  
- "SNP", "CHR", and "BP" as first three columns.
- It has at least one of these columns: "Z","OR","BETA","LOG_ODDS","SIGNED_SUMSTAT"

```{r, eval=FALSE}
path_formatted <- MAGMA.Celltyping::get_example_gwas(
  trait = "prospective_memory")
```

### Map SNPs to Genes

Note you can input the genome build of your summary statistics for this step or
it can be inferred if left `NULL`:

```{r, eval=FALSE}
genesOutPath <- MAGMA.Celltyping::map_snps_to_genes(
  path_formatted = path_formatted,
  genome_build = "GRCh37")
```

### MAGMA_Files_Public

Rather than preprocessing the GWAS yourself, you can instead use the [`MAGMA_Files_Public`](https://github.com/neurogenomics/MAGMA_Files_Public) database we have created. It contains pre-computed MAGMA SNP-to-genes mapping files for hundreds of GWAS.

You can browse which GWAS traits are available by looking at the provided  [*metadata.csv*](https://github.com/neurogenomics/MAGMA_Files_Public/blob/master/metadata.csv) file.

```{r, eval = TRUE}
magma_dirs <- MAGMA.Celltyping::import_magma_files(ids = "ieu-a-298")
```


## CellTypeDataset

`ewceData` provides a number of CellTypeDatasets (CTD) to be used a cell-type
transcriptomic signature reference files.

If you want to create your own single-cell transcriptomic reference, you'll need
to first convert it to CTD using the instructions found in the `EWCE` package
[documentation here](https://github.com/NathanSkene/EWCE/).

```{r} 
ctd <- ewceData::ctd()
```

Note that the cell type dataset loaded in the code above is the Karolinksa cortex/hippocampus data only. For the full Karolinska dataset with hypothalamus and midbrain instead use the following:

```
ctd <- MAGMA.Celltyping::get_ctd("ctd_allKI")
```

Or for the DRONC seq or AIBS datasets use:

```
ctd <- get_ctd("ctd_Tasic")
ctd <- get_ctd("ctd_DivSeq")
ctd <- get_ctd("ctd_AIBS")
ctd <- get_ctd("ctd_DRONC_human")
ctd <- get_ctd("ctd_DRONC_mouse")
ctd <- get_ctd("ctd_BlueLake2018_FrontalCortexOnly")
ctd <- get_ctd("ctd_BlueLake2018_VisualCortexOnly")
ctd <- get_ctd("ctd_Saunders")
```
 
# Run cell-type enrichment analyses  

`MAGMA.Celltyping` offers a suite of functions for conducting various types of cell-type-specific enrichment tests on GWAS summary statistics. 

The `celltype_associations_pipeline` wraps several functions that in previous versions of `MAGMA.Celltyping` had to be set up and run separately. These include: 

- **Linear enrichment**: `calculate_celltype_associations(EnrichmentMode = "linear")` internally. Activated when `run_linear=TRUE`. 
- **Top 10% enrichment**: Uses `calculate_celltype_associations(EnrichmentMode = "Top 10%")` internally. Activated when `run_top10=TRUE`. 
- **Conditional enrichment**: Uses `calculate_conditional_celltype_associations` internally. Activated when `run_conditional=TRUE`.  

Thus, `celltype_associations_pipeline` is designed to make these analyses easier to run. 

```{r, eval = FALSE}
MAGMA_results <- MAGMA.Celltyping::celltype_associations_pipeline(
  magma_dirs = magma_dirs,
  ctd = ctd,
  ctd_species = "mouse", 
  ctd_name = "Zeisel2015", 
  run_linear = TRUE, 
  run_top10 = TRUE)
```

We've also saved a pre-computed version of these results as a dataset:

```{r}
MAGMA_results <- MAGMA.Celltyping::enrichment_results
```

# Plot results

## Merge results  

`merge_results` imports each of the MAGMA enrichment results files and merges them into one so that they can easily be plotted and further analysed. 

```{r}
merged_results <- MAGMA.Celltyping::merge_results(
  MAGMA_results = MAGMA_results)
knitr::kable(merged_results)
```

## Heatmap   

Now we'll construct a heatmap visualizing the enrichment results, such that each GWAS is shown on the y-axis and each cell-type is shown on the x-axis. Results can be further facetted by what kind of test was run (linear, top10%, and/or conditional). 

```{r Heatmap, error=TRUE}  
heat <- MAGMA.Celltyping::results_heatmap(
  merged_results = merged_results, 
  title = "Alzheimer's Disease (ieu-a-298) vs. nervous system cell-types (Zeisel2015)",
  fdr_thresh = 1)
```


# Top results

## Top phenotypes 

Get the phenotypes with the greatest number of significant cell-type enrichment results. 

```{r}
top_phenos <- merged_results %>% 
  dplyr::group_by(EnrichmentMode, GWAS) %>%
  dplyr::summarise(Celltype=dplyr::n_distinct(Celltype)) %>%
  dplyr::arrange(desc(Celltype))
print(top_phenos)
```

## Top enrichments 

Get the phenotypes-celltype enrichment results with the most significant p-values (per phenotype).

```{r} 
top_enrich <- merged_results %>% 
  dplyr::group_by(EnrichmentMode, GWAS) %>%
  dplyr::slice_min(FDR, n = 2)
print(top_enrich) 
```

# Session Info

<details>

```{r}
utils::sessionInfo()
```

</details>
<hr>
